Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-07 _tex/paper2.tex
Built site for gh-pages

Although sampling and comparison mechanisms are central to modern CJ
assessments, it is striking that most of the CJ literature has examined
them only within a limited scope. Researchers have primarily
investigated the effects of adaptive comparative judgment (ACJ) designs
on trait reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}. While
these studies offer valuable insights, they often overlook the broader
role these mechanisms play within the CJ system. As this oversight
likely stems from a broader lack of conceptual clarity about how these
mechanisms function within the system, this study decided to formally
incorporate these mechanisms into CJ's conceptual model.

The explicit integration of the sampling and comparison mechanisms
offers a new perspective on how these mechanisms shape the CJ process.
In particular, it clarifies their role in the data-generating process by
framing them as sources of missing data--that is, as mechanisms that
determine which observations are missing from the final data sample.
This new perspective encourages the application of Little and Rubin's
principled missing data framework \citeyearpar{Little_et_al_2020},
enabling a more rigorous evaluation of existing claims about these
missing data mechanisms, their influence on CJ outcomes, and their
implications for designing and assessing more complex assessments
setups.

@@

However, many modern CJ applications rely on more complex assessment
designs, in which the sampling and comparison mechanisms introduce more
intricate forms of missingness --- such as \emph{missing at random}
(MAR) or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}.
A prominent example is the previously discussed ACJ designs, where prior
judgment outcomes inform the selection of stimulus pairs for subsequent
comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
pair selection process suggests that ACJ's comparison mechanism is
outcome-dependent, potentially classifying the method as a generator of
MNAR data. If this classification holds, the mixed findings on ACJ's
capabilities become more comprehensible: some studies find that the
method improves trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
contend that it artificially inflates these gains
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.

@@

Despite the growing notion that various stimulus-related factors can
influence judges' perceptions
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}
and that these influences may not always cancel each other out, few
studies in the CJ literature provide empirical evidence for judges'
biases
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
This gap likely persists not due to a lack of interest or research but
because researchers often rely on ad-hoc detection methods--such as
`misfit' statistics--that are poorly suited for the task
\citep{Kelly_et_al_2022}. To address this limitation, this study treats
judges' biases as an integral component of the CJ system from the
outset. This approach offers one key advantage: it provides a more
accurate representation of the data-generating process behind pairwise
comparisons, one that acknowledges that the discriminal processes of
stimuli become an observable outcome only through judges' perceptions,
which may exhibit bias.

The explicit integration of judges' bias into CJ's conceptual model then
paves the way for investigating several relevant research questions. One
key question is whether researchers can validly analyze CJ data under
the assumption of ``sample-free'' trait calibration, specifically under
the hypothesis that judges exhibit no systematic bias. This question is
particularly relevant because many researchers still regard
``sample-freeness'' as an inherent property of the BTL model
\citep{Bramley_2008, Andrich_1978} despite growing evidence of
persistent biases. Another critical question, discussed in
Section~\ref{sec-theory-issue1b}, is whether training or expertise can
help judges avoid focusing on irrelevant stimulus features--such as
handwriting--over more central criteria like argumentative quality in
writing assessments \citep{Kelly_et_al_2022}. Notably, exploring these
questions may also provide insights into what it truly means to be an
``expert'' within the CJ context \citep{Kelly_et_al_2022}.

Moreover, since judges draw on a variety of intricate stimulus
characteristics when evaluating complex, multidimensional traits
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022},
and these characteristics account for variation in judgment accuracy
\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020, Gijsen_et_al_2021},
it is reasonable to expect that judges' assessments also vary with
attributes such as gender, age, culture, income, education, training, or
expertise \citep{Kelly_et_al_2022}. Prior studies support this view
\citep{Bartholomew_et_al_2020a, McMahon_et_al_2015}. Thus, building on
the discussion in Section~\ref{sec-discussion_RA1}, researchers could
further explore how judges' selection influences the formation of a
``shared consensus'' and whether these attributes introduce systematic
biases or distortions in the observed trait distribution
\citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
undermine the assumption of ``sample-freeness,'' researchers should
explore ways to control their influence and determine how many
judges--and how many judgments per judge--are needed to produce reliable
trait estimates in these contexts. Additionally, it is worth considering
whether \emph{repeated measures designs}, in which judges evaluate the
same stimulus pairs multiple times \citep{Lawson_2015}, can improve
judgment consistency and accuracy. Our approach provides the necessary
structure to rigorously investigate these questions.

\subsubsection{The identification of `misfitting' stimuli and
judges}\label{sec-discussion_RA3}

Although the CJ literature clearly defines \emph{misfit} judges and
stimuli, researchers have rarely examined how these observations relate
to Thurstonian theory. In particular, they have not identified which
elements of Thurstone's theory might explain the occurrence of misfits.
This disconnect likely stems from the fact that CJ researchers derive
misfit statistics from methods such as residual analysis and outlier
detection rather than from Thurstonian principles
\citep{Pollitt_2012a, Pollitt_2012b, Wu_et_al_2022}. They typically
define \emph{misfit judges} as those whose assessments diverge
significantly from the ``shared consensus''
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
and \emph{misfit stimuli} as those that elicit more judgment
discrepancies than others
\citep{Pollitt_2004, Pollitt_2012a, Pollitt_2012b, Goossens_et_al_2018}.
Both definitions closely mirror the statistical concept of
outliers--that is, observations that deviate markedly from the rest of
the sample in which they occur \citep{Grubbs_1969}. However, this
resemblance extends beyond these definitions. CJ researchers also
identify misfits using standard outlier detection procedures, such as
transforming BTL model residuals into statistics of interest and
comparing them against a predefined threshold.

{After identifying misfit observations, researchers often propose
various hypotheses to explain their presence. Yet, without providing
evidence for any of these explanations, they typically exclude these
observations from analysis. This exclusion process raises three major
concerns: one related to the reliability of the methods used to identify
misfits, a second concerning the potential impact of their removal on
trait estimates and inferences, and a third regarding the hypotheses
used to explain the presence and exclusion of misfits.}

{First, regarding the reliability of identification methods, it is
relevant to note that outlying observations can only be defined relative
to a specific model \citep{McElreath_2020}. Consequently, methods based
on models such as the BTL--which rest on strong assumptions--should be
treated with caution, as these assumptions may prevent the model from
accurately capturing the underlying data-generating process of the CJ
system, thereby potentially compromising the identification of misfits.}

{Second, concerning the impact of removing misfits, the statistical
literature warns that ad-hoc outlier exclusion procedures risk
discarding valuable information \citep{Miller_2023} and introducing bias
into trait estimates. The direction and magnitude of this bias are often
unpredictable, as they depend on which observations researchers exclude
from the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.}

{Lastly, regarding the hypotheses explaining the presence and exclusion
of misfits, it is important to note that a significant number of these
hypotheses could not be tested within the BTL model due to its strong
assumptions and limited flexibility. This is the reason why there is a
lack of supporting evidence for misfit observations. In contrast, our
approach enables researchers to test these hypotheses. For instance,
researchers could examine whether misfit judges are simply those who
exhibit (an outlying degree of) systematic bias or whether they display
greater variability in their judgments compared to their peers.
Similarly, in the case of misfit stimuli, researchers could investigate
whether their status reflects an increased variability in their
discriminal processes or stems from their classification as outlying
stimuli.}

{Moreover, since outliers are not necessarily ``bad data''
\citep{McElreath_2020}, our approach offers a principled alternative
that retains misfits in the analysis without compromising trait
estimation. This alternative involves adapting our proposed models into
more sophisticated, robust measurement models
\citep{Wainer_et_al_1978}--a broad class of procedures designed to
reduce the sensitivity of parameter estimates to mild or moderate
departures from model assumptions \citep{Everitt_et_al_2010}.}

@@

Even after acquiring the necessary foundational knowledge, CJ
researchers face two additional challenges when applying our approach to
their specific research context. First, they must verify whether the
conceptual model provides a \emph{faithful} \footnote{Avoid confusing
  this term with the \emph{faithfulness assumption}
  \citep{Neal_2020, Hernan_et_al_2025} described in the causal inference
  literature. For further details, see footnote 1.} representation of
the CJ system under study. Second, they need to assess whether the
statistical translation of the model can accurately estimate the
intended estimands (i.e., parameters) from empirical data. To ensure a
faithful representation of the CJ system, we encourage researchers to
treat our conceptual and statistical models as starting points rather
than universal solutions for all CJ designs or datasets. While these
models may prove adequate in some contexts, researchers cannot assume
this apriori. Instead, they need to adapt these models to their specific
CJ contexts, paying close attention to assessment design features and
causal assumptions. As discussed in
Section~\ref{sec-discussion_challenges}, our approach facilitates this
process by offering a transparent framework for articulating new
assumptions and guiding the design of CJ assessments.

Conversely, to evaluate the estimation capabilities of the statistical
model, researchers must tackle the challenge of performing
identification analysis. As outlined in Section~\ref{sec-theoretical},
\emph{identification analysis} determines whether a statistical model
can accurately compute a given estimand (e.g., a parameter) based solely
on its (causal) assumptions, independent of random variability
\citep{Schuessler_et_al_2023}. Identification is crucial because it is a
necessary condition for consistency. \emph{Consistency} is the property
of an estimator (e.g., a statistical model) whose estimates converge to
the ``true'' value of an estimand as data size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, consistency is
impossible--even with infinite, error-free data--and thus, meaningful
inference from finite samples cannot be achieved
\citep{Schuessler_et_al_2023}.

Although performing identification analysis through formal derivations
may seem like a natural next step, we argue that the complexity of the
CJ system renders such an approach impractical at the outset. Instead,
simulation-based methods, such as power analysis, provide a more
practical and flexible alternative, enabling researchers to examine the
consistency of estimates without relying on cumbersome mathematical
proofs. Nonetheless, the development of formal identification proofs
remains a significant goal for future CJ research. Importantly, our
approach supports both strategies by providing the probabilistic
foundation for formal derivations and the necessary statistical
structure for simulation-based methods.

@@

\section{Appendix}\label{sec-appendix}

\subsection{Statistical and Causal inference}\label{sec-appendixB}

This section introduces fundamental statistical and causal inference
concepts necessary for understanding the core theoretical principles
described in this document. It does not, however, offer a comprehensive
overview of causal inference methods. Readers seeking more in-depth
understanding may wish to explore introductory papers such as
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find it helpful to consult
introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
\citet{McElreath_2020}. For more advanced study, readers may refer to
seminal intermediate papers such as \citet{Neyman_et_al_1923},
\citet{Rubin_1974}, \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009},
as well as books such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
and \citet{Hernan_et_al_2025}.

\subsubsection{Empirical research and randomized
experiments}\label{sec-appendixB1}

Empirical research uses evidence from observation and experimentation to
address real-world challenges. In this context, researchers typically
formulate their research questions as \emph{estimands} or \emph{targets
of inference}, i.e., the specific quantities they seek to determine
\citep{Everitt_et_al_2010}. For instance, researchers might be
interested in answering the following question: ``To what extent do
different teaching methods \((T)\) influence students' ability to
produce high-quality written texts \((Y)\)?'' To investigate this,
researchers could randomly assign students to two groups, each exposed
to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
perform pairwise comparisons, generating a dichotomous outcome
\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
ability. In this scenario, the research question can be rephrased as the
estimand, ``\emph{On average}, is there a difference in the ability to
produce high-quality written texts between the two groups of students?''
and this estimand can be mathematically represented by the random
associational quantity in Equation~\ref{eq-group_diff}, where
\(E[\cdot]\) denotes the expected value.

\begin{equation}\phantomsection\label{eq-group_diff}{
E[Y_{i} \mid T_{i}=1] - E[Y_{i} \mid T_{i}=2]
}\end{equation}

Researchers then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute the estimand based solely on its assumptions, regardless of
random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
\emph{estimator} refers to a method or function that transforms data
into an estimate \citep{Neal_2020}. \emph{Estimates} are numerical
values that approximate the estimand derived through the process of
\emph{estimation}, which integrates data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020}, shown in Figure~\ref{fig-IEflow},
visually represents the transition from estimands to estimates.

\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{images/png/IEflow.png}

\caption{\label{fig-IEflow}Identification-Estimation flowchart.
Extracted and slightly modified from \citet[pp.~32]{Neal_2020}}

Identification is a necessary condition to ensure \emph{consistent}
estimators. An estimator achieves \emph{consistency} when it converges
to the ``true'' value of an estimand as the data size approaches
infinity \citep{Everitt_et_al_2010}. Without identification, researchers
cannot achieve consistency, even with ``infinite'' and error-free data.
As a result, deriving meaningful insights about an estimand from finite
data becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore,
to ensure accurate and reliable estimates, researchers prioritize
estimators with desirable identification properties. For instance, the
Z-test is a widely used estimator for comparing group proportions,
yielding accurate estimates when its underlying assumptions are
satisfied \citep{Kanji_2006}. Furthermore, researchers can interpret
estimates from the Z-test as causal, provided the data is collected
through a randomized experiment.

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key identification
properties, such as common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from any other variable \(X\)
\citep{Morgan_et_al_2014, Neal_2020}.

Nevertheless, researchers often face constraints that limit their
ability to conduct randomized experiments. These constraints include
ethical concerns, such as the assignment of individuals to potentially
harmful interventions, and practical limitations, such as the
infeasibility of, for example, assigning individuals to genetic
modifications or physical impairments \citep{Neal_2020}. In these cases,
causal inference offers a valuable alternative for generating causal
estimates and understanding the mechanisms underlying specific data. In
addition, the framework can provide significant theoretical insights
that can enhance the design of experimental and observational studies
\citep{McElreath_2020}.

\subsubsection{Identification under causal
inference}\label{sec-appendixB2}

Unlike classical statistical modeling, which focuses primarily on
summarizing data and inferring associations, the \emph{causal inference}
framework is designed to identify causes and estimate their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. The framework uses
rigorous mathematical techniques to address the \emph{fundamental
problem of causality}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This problem
revolves around the question, ``What would have happened `in the world'
under different circumstances?'' This question introduces the concept of
counterfactuals, which are instrumental in defining and identifying
causal effects.

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendixB1}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes, as in Equation~\ref{eq-ICE}.

\begin{equation}\phantomsection\label{eq-ICE}{
\tau_{i} = Y_{i} \mid do(T_{i}=1) - Y_{i} \mid do(T_{i}=2)
}\end{equation}

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \mid do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \mid do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address this challenge, the structural approach extends
the ICE to the \emph{average causal effect} (ACE,
Equation~\ref{eq-ACE}), representing the average difference between the
students' observed potential outcomes and their counterfactual
counterparts.

\begin{equation}\phantomsection\label{eq-ACE}{
\begin{aligned}
\tau & = E[\tau_{i}] \\
  & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)]
\end{aligned}
}\end{equation}

Even though counterfactuals are unobservable, researchers can still
identify the ACE from associational estimates by leveraging the
structural approach. The approach identifies the ACE by statistically
conditioning data on a \emph{sufficient adjustment set} of variables
\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. When such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \mid X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).

Conditioning on a sufficient adjustment set enables researchers to
estimate the ACE, even when the data comes from an observational study.
This process is feasible because such conditioning ensures that the ACE
estimator satisfies several critical properties, including confounding
elimination \citep{Morgan_et_al_2014}. Naturally, the validity of claims
about the causal effects of \(T\) on \(Y\) now hinges on the assumption
that \(X\) serves as a sufficient adjustment set. However, as
\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
real world from observed data inevitably requires assumptions. This
requirement holds true for both observational and experimental data.

For instance, if researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendixB1} and must instead rely on
observational data, they can still identify the ACE as long as an
observed variable \(X\), such as the socio-economic status of the
school, satisfies the backdoor criterion. Under these circumstances,
researchers first identify the \emph{conditional average causal effect}
(CACE, Equation~\ref{eq-CACE})

\begin{equation}\phantomsection\label{eq-CACE}{
CACE_{t} = E[Y_{i} \mid T_{i}=t, X]
}\end{equation}

From the CACE, researchers can identify the ACE from associational
quantities as in Equation~\ref{eq-mACE1}. This identification process is
commonly known as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\)
represents the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}.

\begin{equation}\phantomsection\label{eq-mACE1}{
\begin{aligned}
  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
  & = E_{X}[CACE_{1} - CACE_{2}] \\
  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

Notably, the approach extends the ACE identification for a continuous
variable \(T\) as in Equation~\ref{eq-mACE_cont}, ensuring broad
applicability across different causal scenarios
\citep[pp.~45]{Neal_2020}

\begin{equation}\phantomsection\label{eq-mACE_cont}{
\begin{aligned}
  \tau &= E[Y_{i} \mid do(T_{i}=t)] \\
  & = d E_{X}\left[ E[Y_{i} \mid T_{i}=t, X]\right]/ dt
  \end{aligned}
}\end{equation}

\subsubsection{Diving into the specifics}\label{sec-appendixB3}

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, these tools serve as \emph{conceptual (theoretical) models}
on which identification analysis rests
\citep[pp.~4]{Schuessler_et_al_2023}. Thus, using these tools,
researchers can determine which statistical models can identify (ACE,
CACE, or other), assuming the depicted causal structure is correct
\citep{McElreath_2020}, thus enabling valid causal inference.
Figure~\ref{fig-IEflow} shows the role of theoretical models in the
inference process.

SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.

Thus, Section~\ref{sec-appendixB31} and Section~\ref{sec-appendixB32}
elaborate on the first advantage, while Section~\ref{sec-appendixB32}
and Section~\ref{sec-appendixB33} do so for the second. Finally,
Section~\ref{sec-appendixB34} explains how researchers use SCMs and DAGs
alongside Bayesian inference methods in the estimation process.

\paragraph{The five fundamental block for SCMs and
DAGs}\label{sec-appendixB31}

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display the five fundamental building blocks for SCMs and DAGs. The left
panels of the figures show the formal mathematical models, represented
by the SCMs, defined in terms of a set of \emph{endogenous} variables
\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

>>>>

Although sampling and comparison mechanisms are central to modern CJ
assessments, it is striking that most CJ literature has examined them
within a limited scope. Researchers have primarily investigated the
effects of adaptive comparative judgment (ACJ) designs on trait
reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}. While
these studies offer valuable insights, they also overlook the broader
role that these mechanisms play within the CJ system. As this oversight
likely stems from a more fundamental lack of conceptual clarity about
how these mechanisms function within the system, this study decided to
integrate these mechanisms into the conceptual model of CJ.

The explicit integration of the sampling and comparison mechanisms
offers a new perspective on how these mechanisms shape the CJ process.
Specifically, it clarifies their role as sources of missing data in CJ's
data-generating process--that is, as mechanisms that determine which
observations are missing from the final data sample. This new
perspective encourages the application of Little and Rubin's principled
missing data framework \citeyearpar{Little_et_al_2020}, allowing a more
rigorous evaluation of existing claims about these missing data
mechanisms, their influence on CJ outcomes, and their implications for
designing and evaluating more complex assessments setups.

@@

However, many modern CJ applications rely on more complex assessment
designs, in which the sampling and comparison mechanisms introduce more
intricate forms of missingness--such as \emph{missing at random} (MAR)
or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
prominent example is the previously discussed ACJ designs, where prior
judgment outcomes inform the selection of stimulus pairs for subsequent
comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
pair selection process suggests that ACJ's comparison mechanism is
outcome-dependent, potentially classifying the method as a generator of
MNAR data. If this classification holds, the mixed findings on ACJ's
capabilities become more comprehensible: some studies find that the
method improves trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
contend that it artificially inflates these gains
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.

@@

Despite the growing notion that various stimulus-related factors
influence judges' perceptions
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}
and that these influences may not always cancel each other out, few
studies in the CJ literature provide empirical evidence for judges'
biases
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
This gap likely persists not due to a lack of interest or research but
because researchers often rely on ad-hoc detection methods, such as
`misfit' statistics, that may not be well-suited for the task
\citep{Kelly_et_al_2022}. To address this limitation, this study treats
judges' biases as an integral component of the CJ system from the
outset. This approach offers one key advantage: it provides a more
accurate representation of the data-generating process behind pairwise
comparisons, one that acknowledges that the discriminal processes of
stimuli become an observable outcome only through judges' perceptions,
which may exhibit bias.

The explicit integration of judges' bias into CJ's conceptual model then
paves the way for investigating several relevant research questions. One
key question is whether researchers can validly analyze CJ data under
the assumption of ``sample-free'' trait calibration, specifically under
the hypothesis that judges exhibit no systematic bias. This question is
particularly relevant because many researchers still regard
``sample-freeness'' as an inherent property of the BTL model
\citep{Bramley_2008, Andrich_1978} despite growing evidence of
persistent biases. Another critical question, discussed in
Section~\ref{sec-theory-issue1b}, is whether training or expertise can
help judges avoid focusing on irrelevant stimulus features, such as
handwriting, over more central criteria like argumentative quality in
writing assessments \citep{Kelly_et_al_2022}. Notably, exploring these
questions may also provide insights into what it truly means to be an
``expert'' within the CJ context \citep{Kelly_et_al_2022}.

Moreover, since judges rely on these stimulus-related factors when
evaluating complex, multidimensional traits
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022},
and these factors account for variation in judgment accuracy
\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020, Gijsen_et_al_2021},
it is reasonable to expect that assessments also vary according to
judge-specific attributes such as gender, age, culture, income,
education, training, or expertise \citep{Kelly_et_al_2022}. Prior
studies support this view
\citep{Bartholomew_et_al_2020a, McMahon_et_al_2015}. Thus, building on
the discussion in Section~\ref{sec-discussion_RA1}, researchers could
further explore how judges' selection influences the formation of a
``shared consensus'' and whether these attributes introduce systematic
biases or distortions in the observed trait distribution
\citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
undermine the assumption of ``sample-freeness,'' it becomes essential to
explore strategies for mitigating their influence and to determine how
many judges (and how many judgments per judge) are needed to produce
reliable trait estimates under these conditions. In addition, it is
worth considering whether \emph{repeated measures designs}, in which
judges evaluate the same stimulus pairs multiple times
\citep{Lawson_2015}, can improve judgment consistency and accuracy. As
anticipated, our approach provides the necessary structure to
investigate rigorously these questions.

\subsubsection{The identification of `misfitting' judges and
stimuli}\label{sec-discussion_RA3}

Although the CJ literature clearly defines \emph{misfit} judges and
stimuli, CJ researchers have rarely examined how these observations
relate to Thurstonian theory. In particular, they have not identified
which elements of Thurstone's theory might account for the occurrence of
misfits. This disconnect likely stems from the fact that CJ researchers
derive misfit statistics from residual analysis and outlier detection
methods rather than from Thurstonian principles. Specifically,
\emph{misfit judges} are typically defined as those whose assessments
diverge significantly from the ``shared consensus''
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
while \emph{misfit stimuli} are those that elicit more judgment
discrepancies than others
\citep{Pollitt_2004, Pollitt_2012a, Pollitt_2012b, Goossens_et_al_2018}.
Both definitions closely mirror the statistical concept of
outliers--that is, observations that deviate markedly from the rest of
the sample in which they occur \citep{Grubbs_1969}. But this resemblance
extends beyond the definitions themselves, as CJ researchers often
identify misfits using conventional outlier detection procedures, such
as transforming BTL model residuals into diagnostic statistics and
comparing them against predefined thresholds
\citep{Pollitt_2012a, Pollitt_2012b, Wu_et_al_2022}.

Nevertheless, it is not the classification of misfits as outliers that
raises concerns for trait measurement and inference, but rather the
prevailing CJ practice of identifying them through these ad-hoc
detection procedures and then excluding them from analysis, often
without providing empirical evidence for the various hypotheses proposed
to explain their occurrence in the first place. In this regard, it is
essential to recognize that outliers can only be defined relative to a
specific model \citep{McElreath_2020}. Thus, identification procedures
based on models like the BTL, which relies on strong assumptions, should
be approached with caution, as they may not be well-suited for the task
\citep{Kelly_et_al_2022}. Specifically, because the BTL model may not
always accurately reflect the underlying data-generating process of the
CJ system (see Section~\ref{sec-theory-issues}), detection procedures
grounded in this model risk compromising the accurate identification of
misfits. Moreover, the statistical literature cautions that excluding
outliers risks discarding valuable information \citep{Miller_2023} and
introducing bias into trait estimates. The direction and magnitude of
this bias are often unpredictable, as they depend on which observations
researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. Lastly, regarding
the hypotheses proposed to explain the presence of misfits, it is
evident that the BTL model lacks the flexibility needed to adequately
test many of them, largely due to its strong assumptions.

In contrast, our approach provides a rigorous framework that enables the
examination of several relevant hypotheses. For instance, researchers
could investigate whether misfit judges are those who exhibit (an
outlying degree of) systematic bias or greater variability in their
judgments compared to their peers. Similarly, researchers might explore
whether misfit stimuli exhibit more variable discriminal processes
relative to other stimuli or if they are genuinely outlying cases.
Moreover, since outliers are not necessarily ``bad data''
\citep{McElreath_2020}, our approach also offers a principled
alternative that retains misfits in the analysis without compromising
trait estimation or inference. This alternative involves adapting our
proposed models into robust measurement models \citep{McElreath_2020}--a
broad class of procedures designed to reduce the sensitivity of
parameter estimates to mild or moderate departures from model
assumptions \citep{Everitt_et_al_2010}. This alternative also relates to
a broader discussion in social science, which is: that when researchers
face low predictive capacity from their models, they should not only
search for ``new'' variables or alternative procedures but also consider
employing more sophisticated measurement models
\citep{Wainer_et_al_1978}.

@@

Even after acquiring the necessary foundational knowledge, CJ
researchers still face two additional challenges when applying our
approach to their specific research context. First, they must verify
whether the conceptual model provides a \emph{faithful} \footnote{Avoid
  confusing this term with the \emph{faithfulness assumption}
  \citep{Neal_2020, Hernan_et_al_2025} described in the causal inference
  literature. For further details, see footnote 1.} representation of
the CJ system under study. Second, they need to assess whether the
statistical translation of the model can accurately estimate the
intended estimands (i.e., parameters) from empirical data. To ensure a
faithful representation of the CJ system, we encourage researchers to
treat our conceptual and statistical models as starting points rather
than universal solutions for all CJ designs or datasets. While these
models may prove adequate in some contexts, researchers cannot assume
this apriori. Instead, they need to adapt these models to their specific
contexts, paying close attention to assessment design features and
causal assumptions. As discussed in
Section~\ref{sec-discussion_challenges}, our approach facilitates this
process by offering a transparent framework for articulating new
assumptions and guiding the design of CJ assessments.

Conversely, to evaluate the estimation capabilities of the statistical
model, researchers must tackle the challenge of performing
identification analysis. As outlined in Section~\ref{sec-theoretical},
\emph{identification analysis} determines whether a statistical model
can accurately compute a given estimand (e.g., a parameter) based solely
on its (causal) assumptions, independent of random variability
\citep{Schuessler_et_al_2023}. Identification is crucial because it is a
necessary condition for consistency. \emph{Consistency} is the property
of an estimator (e.g., a statistical model) whose estimates converge to
the ``true'' value of an estimand as data size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, consistency is
impossible--even with infinite, error-free data--and thus, meaningful
inference from finite samples cannot be achieved
\citep{Schuessler_et_al_2023}. Although performing identification
analysis through formal derivations may seem like a natural next step,
the complexity of the CJ system renders such an approach impractical at
the outset. Instead, simulation-based methods, such as power analysis,
provide a more practical and flexible alternative, enabling researchers
to examine the consistency of estimates without relying on cumbersome
mathematical proofs. Nonetheless, the development of formal
identification proofs remains a significant goal for future CJ research.
Importantly, our approach supports both strategies by providing the
probabilistic foundation for formal derivations and the necessary
statistical structure for simulation-based methods.

@@
